---
title: "Dates, Times, TimeIndexes and Time Series"
author: "Jeff Hallman"
date: '2016-11-18'
output:
  html_document:
    toc: no
  github_document:
    toc: no
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, comment="OUTPUT> ")
# source(".Rprofile")
```

# Dates and Times in Base R
## Dates

Early versions of R had a simple Date class, and code that handled somewhat more
general time objects was in a separate `chron` package.  Neither worked very well
for doing various kinds of date arithmetic or as indexes into time seriesm and
this led to a number of people writing their own time and date packages. But
base R time and date code has improved a lot over the years.

The base `Date` class supports dates without times. Not including times simplifies
things a lot, since you no longer have to worry about time zones and daylight
savings times, or what to do if only a date was specified but not a time. Should
the default time be noon? Midnight? 

Here are some Date examples:
```{r Date1}
today <- Sys.Date(); today
today - 1 ## yesterday
today + 7
```
Internally, Dates are represented as the number of days since January 1, 1970.
You can see this by `unclass`ing them:
```{r Date2}
unclass(today)
baseDate <- as.Date("1970-01-01"); unclass(baseDate)
as.integer(today - baseDate)
```
Now you're wondering why the `as.integer` wrapper on that last one. That's
because without it, you get a `difftime` object:
```{r difftime1}
myDifftime <- today - baseDate; myDifftime
unclass(myDifftime)
```
A difftime object represents a time interval. It's a number with a "units"
attribute. You can also create a difftime object explicitly:
```{r difftime2}
twoWeeks <- as.difftime(2, units = "weeks")
today + twoWeeks
```
One problem with difftime's is that the units have to be "linear", i.e., weeks,
days, hours, seconds, etc. Months, quarters and years don't work. If I tell you
to add one month to a date, how many days do you add? It depends on what month
the original date is in. 

## Times (w/Dates)
The POSIX standards specify a system for describing instants in time. R uses two
timeDate classes, `POSIXct` and `POSIXlt` to implement this system. A POSIXct
represents a time as the number of seconds since midnight on January 1, 1970, in
the GMT time zone:
```{r POSIXt1}
aTime <- Sys.time(); aTime
class(aTime)
unclass(aTime)
unclass(as.POSIXct("19700101:000001", tz= "GMT", format = "%Y%m%d:%H%M%S"))
```
The `POSIXlt` represents instants in time using a list with named elements:
```{r POSIXt2}
unclass(as.POSIXlt(Sys.time()))
```
## Date formats
You can convert most date and dateTime objects to strings and back using format
specifiers. For example:
```{r dateFormats}
format(Sys.time(), format = "%Y is year, %m is month, %d is day")
as.character(Sys.time(), format = "%b is abbreviated month name, %H is hour")
format(Sys.time(), format = "There are also compound formats like %c")
as.POSIXct("2016-12-24 23:59", format = "%Y-%m-%d %H:%M") ## Santa arrival time
```
Finally, some generic functions like `seq` have methods that work with
both Dates and POSIXt times:  
```{r seq.POSIXt}
seq(from = Sys.time(), by = "2 hours", length = 8)
```
# TimeIndexes in the `tis` package
The `tis` package defines the `jul` (Julian date) and `ti` (Time Index) classes. 
```{r getlib}
library(tis)
```
## Julian dates
are represented by instances of the `jul` class. Two Julian dates differ by the
number of days between them. This is also true of `Date` objects in base R, but
`jul` dates use a different base date. You can create a `jul` from a `ti` with
the `jul()` function, which like `ymd()` has an optional `offset` argument. Like
`ti's`, `jul` objects can be converted to and created from several other kinds
of date objects.
```{r jul}
dayInYear <- jul(today()) - jul(20151231); dayInYear
```
## ti objects
Time index objects are used for date calculations and as indexes for `tis` time
series. A time index has two parts: a frequency and period. The period
represents the number of periods elapsed since the base period for that
frequency. Adding or subtracting an integer to a time index adds or subtracts
that number to the period part, so that if `z` is a time index representing the
week ending Monday, April 3 2006, `z + 1` is a time index representing the week
ending April 10, `z + 3` represents the week ending April 24, and so on.  

A time index has class 'ti', and is usually created by the `ti` function. 
However, there is a `today()` function that returns a "daily" `ti`. If z
is a ti, you can access its frequency code with `tif(z)` or its period with
`period(z)`. You can also get the frequency name from `tifName(z)`. Note that you
can have a vector `ti`, that is, a vector with class 'ti' whose individual
elements each represent a different time index. Here are some examples:
```{r ti 1}
march2007 <- ti(20070331, "monthly"); march2007
today()
thisWeek <- ti(latestMonday(), "wmonday"); thisWeek
tif(today())
tifName(today())
period(today())
# How this is actually encoded is pretty simple:
tif(thisWeek)
period(thisWeek)
format(unclass(thisWeek), digits = 12)
period(ti(latestMonday() + 28, "wmonday"))
# yyyymmdd dates for the last day of the next 12 months:
ymd(currentMonth() + 0:11)
# the third Tuesday of the year for each of the next 5 years:
ymd(ti(firstDayOf(currentYear() + 1:5), "wtuesday") + 2)
```
If you want to convert a `ti` object to a yyyymmdd date, use the `ymd()` function:
```{r ti 2}
dec57 <- ti(19571231, tif = "monthly")
# By default, ymd(aTi) gives you the last day of the period. To get the first day of the period, use this trick
ymd(jul(dec57 - 1) + 1)
```
The `ti` function needs to know what the frequency of the ti it's creating is
supposed to be, and the first argument has to be a itself a ti object or
something that specifies a point in time. You can supply additional arguments to
help it figure out how to interpret the first argument, e.g.,
```{r ti 3}
ti("3/31/1957", tif = "daily", format = "%m/%d/%Y") + 1
ti("3/31/1957", freq = 12,     format = "%m/%d/%Y") + 1
```
Finally, there are some convenience function like `holidays()` and 
`nextBusinessDay()` which knows about holidays. 

# Time Series
Base R contains a lot of code for representing and analyzing time series data.
The fundamental class is `ts`, which can represent regularly spaced time series,
such as annual, monthly and quarterly data. It doesn't work well with weekly
data because not all years have the same number of weeks. It depends on what day
of the year a particular year starts on and what day of the week the weeks
you're looking at end on. 

The inability of early versions of R to deal well with weekly, biweekly, daily and
business-day data was the impetus behind the predecessors of the `tis` package.
`tis` series have observation times tjat are represented by `ti` sequences. A
`tis` is actually just a vector or matrix with class 'tis' and a `start`
attribute that is the `ti` for the first observation. Then `start - 1 + k` is the
time index for the k'th observation, and so on. If `x` is a `tis`, you can
reference the fourth observation using either the number '4' or the `ti` given
by `(start(x) + 3)` as an index into `x`. Here are some examples of creating a
`tis` series and indexing into it:
```{r tis 1}
x <- tis(1:120, start = latestJanuary() - 108); x
k = 2; start(x) + k
x[start(x) + 3]
x[end(x) + 1] <- x[end(x)]; x
x[end(x) + 3] <- 42; x
```
As you might expect, arithmetic operations (addition, subtraction,
multiplication and division) can all be performed with combinations of scalars
and series, and the 'tis' machinery will insure that everything is lined up and
windowed, and that the return values will also be time indexed series. 
Furthermore, many of the standard R matrix and time series functions have 'tis'
versions as well. For example, if `X` is a multivariate time indexed series with a
column per component, then `rowSums(X)` is a univariate tis, as is `rowMeans(X)`.
The `cbind()` function can be used to add columns to a tis, while `mergeSeries()` can do
what it says. The `lag` and `diff` functions operate as you might expect, and you
can `window` a series to cut off observations before and/or after particular time
indexes.  

## Plotting tis series

### Download Starbux monthly return date into a tis series
Let us get started by downloading the monthly return data from <a
href="http://assets.datacamp.com/course/compfin/sbuxPrices.csv">sbuxPrices.csv</a>,
and by using tisFromCsv() to read a tis series from it.
```{r downloadfile }
# Download the CSV file and look at the first few lines:
url <- "http://assets.datacamp.com/course/compfin/sbuxPrices.csv"
download.file(url, dest = "sbux.csv")
flines <- readLines("sbux.csv"); flines[1:5]
```
There is a problem with the dates. All of the dates except the first one are the
first weekday of a month. Let's make the heroic assumption that the first date
was supposed to be the first of March 1993 rather than the 31'st. Here's a quick
fix in R: 

```{r fixfile }
flines[2] <- gsub("3/31", "3/1", flines[2])
cat(flines, file = "sbux.csv", sep = "\n")
flines <- readLines("sbux.csv"); flines[1:5]
```

Now that looks OK. We'll use the `tisFromCsv` function to read the data into a
`tis` time series. Note we set the stringsAsFactors global option to FALSE to
avoid a PITA

```{r tisFromCsv 1}
options(stringsAsFactors = FALSE)
sbux <- tisFromCsv("sbux.csv", dateCol = "Date", dateFormat = "%m/%d/%Y")
str(sbux)
# sbux is a list of length one, because tisFromCsv returns a list of the series
# it reads in. Let's just make it be the series itself:
sbux <- sbux[[1]]
### Info about sbux
class(sbux)
start(sbux)
frequency(sbux)
tifName(sbux)
dateRange(sbux)
head(ti(sbux))
head(time(sbux))
```
Notice that `time(x)` returns the times of the observations in years, with the
first day of the year counting as day 0 of that year. So `time(jul(20010101))` is
2001, while `time(jul(20001231))` is 2000.997.

To get the subseries that starts and ends on particular dates, use the `window()` function:
```{r window}
sbuxShort <- window(sbux, start = 19940301, end = 19950301)
sbuxShort
```
What happened here is that the `window` method for `tis` objects helpfully first
converted the `start` and `end` parameters to `ti` objects with the same `tif`
as `sbux`, and then cut off the series before the start ti and after the end ti. 

### Subsetting directly
If you just want the observation for July 1996, you can do this:
```{r accessByTi}
sbux[ti(19960701, tif = tif(sbux))]
# Verify this (note the c(year, period) form for start and end
window(sbux, start = c(1996, 1), end = c(1996,12))
```
If we want to plot the data, we need some X-axis coordinates. One way to do this
is with the `time()` function we saw above:
```{r plot}
plot(time(sbux), sbux)
# or dress it up a little
plot(time(sbux), sbux, type = "l", col = "blue", 
     lwd = 2, ylab = "Adjusted close",
     main = "Monthly closing price of SBUX")
grid()
```

or we can use a function that knows something about tis series:
```{r tisPlot}
tisPlot(sbux)
# and dress this one up a bit
tisPlot(sbux, color = "green", 
        xTickFreq = "annual", xTickSkip = 2,
        xUnlabeledTickFreq = "annual",
        xMinorTickFreq = "quarterly",
        head = "Monthly closing price of SBUX")
```


