---
title: "Lecture 11, dplyr"
date: "2016-12-02"
output: 
#   pdf_document:
#     toc: yes
#     toc_depth: 3
#  word_document:
#    fig_width: 5
#    fig_height: 5
#   fig_caption: true
  html_document:
    highlight: pygments
    theme: spacelab
    toc: yes
header-includes: \usepackage{graphicx}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)#, comment="OUTPUT> ", background='yellow' )
options("width"=512)
```

## Topics 

In this lecture we will cover some of the main features of dplyr, a popular package for manipulating data frames. The package is Hadley Wickham's re-imagined plyr package (with underlying C++ secret sauce co-written by Romain Francois). plyr 2.0 if you will. It does less than plyr, but what it does it does more elegantly and much more quickly. 

dplyr provides: 

* a set of functions for efficiently manipulating data sets, and
* a grammar of data manipulation

Let's load the package.

```{r load_package}
library(dplyr)
help(dplyr)
```


## Example data

We will use these data sets to demonstrate the use of the dplyr tools
```{r get_batting_data}
library(Lahman)
help(Lahman)
```

Each row in the Master data set represents one player
```{r str_master}
str(Master)
```

Each row in Batting represents one year of data for one player 
```{r str_batting}
str(Batting)
```


Each row in Batting represents one year of data for one player 
```{r str_salaries}
str(Salaries)
```


## dplyr verbs

dplyr is built around 5 main functions. 

Each function takes a data set as input, and returns a data set as output. Each function name is a "verb" that describes the specific action being performed on the data set 

These verbs cover the most common data manipulation tasks

* Select certain columns of data.
* Filter your data to select specific rows.
* Arrange the rows of your data into an order.
* Mutate your data frame to contain new columns.
* Summarize chunks of you data in some way.

Let's look at how those work.


## Verb 1: Select

Include or exclude specific variables

The master table has `r ncol(Master)` columns
```{r }
dim(Master)
```

Include only a few variables by listing their names. Do not quote the variable names.

```{r select_example1}
df <- select(Master, playerID, nameFirst, nameLast, birthYear, birthCountry, weight, height, bats, throws)

dim(df)
head(df)
```

Exclude unwanted variables using a minus sign
```{r select_example2}
df <- select(df, -playerID, -birthYear, -birthCountry)

dim(df)
head(df)
```

## Verb 2: Filter 

Include or exclude specific observations

The master table has `r nrow(Master)` rows
```{r }
dim(Master)
```

Include only a few rows by writing a condition. Again do not quote the variable names.

```{r filter_example1}
df <- select(Master, playerID, nameFirst, nameLast, birthYear, birthCountry, weight, height, bats, throws)

dim(df)

df <- filter(df, birthYear >= 1990 & bats == "L")

dim(df)
head(df)
```



## Verb 3: Arrange 

Change the order of the observations

The Salaries data set comes sorted by year, team, and playerID
```{r }
head(Salaries)
```

Sort observations in ascending order by salary and year
```{r sort_ascending}
df <- arrange(Salaries, salary, yearID)
head(df)
```


Sort observations in descending order by salary and year
```{r sort_desceding}
df <- arrange(Salaries, -salary, -yearID)
head(df)
```


## Verb 4:  Mutate 

Add new columns to a data set. Create a new value for every row

Compute one new variable: runs per game
```{r mutate_batting}
df <- select(Batting, playerID, yearID, R, G)
dim(df)

df <- mutate(df, RPG = R/G)
dim(df)
head(df)
```


Add several new variables

* runs per game for each player
* home runs per game
* bases stolen per game

And, change an existing variable

* Make Runs negative
```{r mutate_batting2}
df <- select(Batting, playerID, yearID, HR, SB, R, G)
dim(df)

df <- mutate(df, 
                 RPG  = R/G,
                 HRPG = HR/G,
                 SBPG = SB/G,
                 R    = -1 * R)
dim(df)
head(df)
```

Notice that the change to the Runs variable took effect last, otherwise the other variables we created would have been negative.

## Verb 5:  Summarise 

 Create a new variable using data from multiple observations, reduce the number observations

Compute average height of players
```{r summarize1}
df <- select(Master, height, weight)
dim(df)

df <- summarise(df, avg_height = mean(height, na.rm = TRUE))
dim(df)

print(df)
```

Compute average height (in feet) and weight of players, along with standard deviation
```{r summarize2}
df <- select(Master, height, weight)
dim(df)

df <- summarise(df, 
                   avg_height = mean(height, na.rm = TRUE) / 12,
                    sd_height = sd(height, na.rm = TRUE),
                   avg_weight = mean(weight, na.rm = TRUE),
                    sd_weight = sd(weight, na.rm = TRUE))
dim(df)
print(df)
```


## Pipes
dplyr provides another innovation over plyr: the ability to chain operations together in sequence with the pipe (%>%) operator

The following examples are equivalent.

### Example Without pipes

Repeatedly input and output a data frame for each step
```{r dplyr_no_chain1}
df <- select(Master, birthYear, bats, height, weight)
df <- filter(df, birthYear >= 1990 & bats == "L")
df <- summarise(df, avg_height = mean(height, na.rm = TRUE))
print(df)
```

### Example With pipes

The output from each step is used as the input for the next step
```{r dplyr_with_chain1}
Master %>% 
    select(birthYear, bats, height, weight) %>%
    filter(birthYear >= 1990 & bats == "L") %>%
    summarise(avg_height = mean(height, na.rm = TRUE))
```

### Pipes explained

The "." is a special object that represents the output from the prior step

```{r dplyr_with_chain2}
Master %>% 
    select(., birthYear, bats, height, weight) %>%
    filter(., birthYear >= 1990 & bats == "L") %>%
    summarise(., avg_height = mean(height, na.rm = TRUE)) 
```


Pipes replace nested function calls -- without them this would be the only way to avoid explicitly creating a temporary data set for each step. 
```{r dplyr_with_chain3}
summarise( 
          filter( 
                 select(
                        Master, 
                        birthYear, bats, height, weight), 
                 birthYear >= 1990 & bats == "L"), 
          avg_height = mean(height, na.rm = TRUE) 
        )
```


Now that we've seen how pipes work, we can use multiple "verbs" to get more useful results


## Aggregate data

*group_by* allows you to summarize data separately for subgroups of observations.

Aggregate by batter type: left handed, right handed, both, and unknown hitters

```{r dplyr_groupby}
Master %>%
  group_by(bats) %>%
  summarise(player_count = n(),
              avg_height = mean(height, na.rm = TRUE) / 12,
              avg_weight = mean(weight, na.rm = TRUE))

```

Aggregate by year

```{r dplyr_groupby2}
Salaries %>%
  filter(yearID >= 2000) %>%
  group_by(yearID) %>%
  summarize(avg_salary = mean(salary, na.rm = TRUE),
             sd_salary = sd(salary, na.rm = TRUE),
            lowest = min(salary),
            highest = max(salary)) %>%
  data.frame()

```

## How do I save output from a pipe?

Assign it to an object 
```{r dplyr_groupby_assign}
df <- Master %>%
          group_by(bats) %>%
          summarise(player_count = n(),
                      avg_height = mean(height, na.rm = TRUE) / 12,
                      avg_weight = mean(weight, na.rm = TRUE)) 

print(df)
```

## Joins

Let's look at players born in 1990 or later and compare the salaries of left vs right-handed hitters

```{r join_example}
players <- Master %>% 
               select(playerID, nameFirst, nameLast, birthYear, bats) %>%
               filter(birthYear >= 1990)

pay <- Salaries %>%
           select(playerID, yearID, salary) %>%
           mutate(salary = round(salary / 1000000, digits = 2)) %>%
           group_by(playerID) %>%
           summarize(avg_salary = mean(salary, na.rm= TRUE),
                     min_salary = min(salary),
                     max_salary = max(salary))
               
dim(players)
dim(pay)           

df <- left_join(players, pay, by = "playerID") 
dim(df)
head(df)

left_join(players, pay, by = "playerID") %>%
    group_by(bats) %>%
    summarize(avg_high_salary = mean(max_salary, na.rm= TRUE))
```


## Other functions

dplyr also provides a function glimpse() that makes it easy to look at our data in a transposed view. It's similar to the str() (structure) function, but has a few advantages (see ?glimpse).

```{r datalook}
glimpse(Salaries)
```


## Conclusion

dplyr makes easier to write clear working code--which allows you to focus on the details of your data analysis.


## For more info

#### One-page cheatsheet

https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf

#### Explanations from creator
http://r4ds.had.co.nz/transform.html

#### Free online training
https://www.datacamp.com/courses/dplyr-data-manipulation-r-tutorial


