---
title: "Lecture 6 - Continued"
author: "Jeffrey Naber"
date: "October 13, 2016"
output: html_document
---

# Dates, Better than Brute Force Techniques


## Dates in R

R treats Dates as a unique data class, represented as:  "YYYY-MM-DD".  R can do most operations (that makes sense) on dates, simply as part of the basic R package.
```{r}
dates <- seq(as.Date("2015-01-01"), to = as.Date("2015-12-31"), by = "month")

mean(dates)

min(dates)

max(dates)
```
R can convert many, many different strings into a date class using the "format" argument in as.Date:

```{r}
as.Date("1980-02-10")

as.Date("19800210", format = "%Y%m%d")

as.Date("10Feb80", format = "%d%b%y")

as.Date("9/18/2016", format = "%m/%d/%Y")
```

Symbol  | Meaning
------- | ---------------------------------
%d      | Day as a number
%a      | Abbreviated Weekday
%A      | Unabbreviated Weekday
%m      | Month as a number
%b      | Abbreviated month
%B      | Unabbreviated month
%y      | Two-digit year
%Y      | Four-digit year

Format is both an argument and a function itself:

```{r}
Sys.Date()
format(Sys.Date(), "%d %B %Y")
format(Sys.Date(), "%d-%b")
format(Sys.Date(), "%Y")

paste0("carsDataSet_", format(Sys.Date(), "%Y%m%d"), ".csv")

```
R can even import from Excel, with Excel's wierd 5-digit dates.
```{r}
# From Windows Excel
as.Date(30829, origin = "1899-12-30")

# From iOS Excel
as.Date(29367, origin = "1904-01-01")
```

Rather technical note:  In the case of two digit years, R (currently) assumes that years 00-68 are 2000 - 2068, and years 69-99 are 1969 - 1999.  

### In Class Exercise:  
Write a short function that reads in a date with a 2-digit year and returns just dates in the 1900s.

## Brute Force as a Last Resort

For loops are very useful, but they're the R equivalent of the Quadratic equation:  Yes, they work, but there's often a much, much better way to solve the problem.

```{r echo = FALSE}
# Create a data frame
exampleData <- data.frame(date = seq(as.Date("2010-01-01"), Sys.Date(), by = "month"))
exampleData$c1 <- runif(nrow(exampleData), 0, 500)
exampleData$c2 <- runif(nrow(exampleData), 200, 700)
exampleData$c3 <- runif(nrow(exampleData), 100, 300)
```

If we wanted to build a dataframe of yearly averages, we could use a for loop:
```{r}
head(exampleData)

exampleData$year <-format(exampleData$date, "%Y")
yearlyMeans <- data.frame(year = unique(exampleData$year), mean = NA)

for(i in unique(exampleData$year)) {
  exampleDataSubset <- exampleData[exampleData$year == i, ]
  yearlyMeans$mean[which(yearlyMeans$year == i)] <- mean(exampleDataSubset$c1)
}

yearlyMeans
```

While this way works, it's bulky, hard to follow, and, particularly once larger dataset get involved, is memory and time intensive.

```{r}
yearlyMeansAggregate <-
  aggregate(exampleData$c1, by = list(format(exampleData$date, "%Y")), FUN = mean)

yearlyMeansAggregate

```

One line, easy to read, and much, much faster, especially with larger datasets.

You'll run into a lot of problems when you are writing code that will look like they can be solved with a for loop.  Take a few minutes to explore StackOverflow, r-bloggers.com, and simple google searches to see if there's a function that can work instead of a loop.
